---
title: "Interactive Graphs with Plotly and Dash"
author: "Giorgio Coppola, Gayatri Shejwal, Lonny Chen"
subtitle: IDS Workshop on interactive data visualization Plotly and Dash in R
output:
  html_document:
    toc: yes
    df_print: paged
    number_sections: no
    highlight: tango
    theme: sandstone
    toc_depth: 3
    toc_float: yes
    css: custom.css
    self_contained: no
  word_document:
    toc: yes
    toc_depth: '3'
institute: Hertie School
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

<br>

Data visualization is a powerful mean to understand complex data and communicate insights effectively, revealing the hidden patterns and trends that wouldn't be obvious in tabular data, turning raw data into information.

The tools we are going to introduce today serve exactly this purpose, enhancing the visualization possibilities offered by other static visualization packages such as `ggplot2`. 


**What tools allow to do so?**

Plotly and Dash! And we will introduce both!

Indeed, `plotly` and `dash` are interactive, in the sense that the user can interact with the data through the graph directly. This allows possibilities for higher understanding of the data. 

### ***Why interactive graphs?

* *Engagement*: Makes data come alive (interactive) and is more engaging for users. This increase understanding and retention.

* *Depth*: Allows users to drill down and explore nuances, as interaction allows to create narratives more easily. Users can ask 'why' and get answers instantly by exploring the chart or graph.

* *Decision-making*: Facilitates more informed decisions.

Both `plotly` and `dash` are interactive, but we can say that the level of interaction is different: `plotly` appears very similar to a static plot, with the difference that we can hover your cursor over it, accessing further data. `dash` on the other hand, creates proper dashboards, namely HTML apps, allowing even more interaction.


# Plotly
### What is Plotly?

Plotly is actually a company headquartered in Montreal, Canada, providing online graphing, analytics, and statistics tools, as well as scientific graphing libraries for Python, R, MATLAB, Perl, Julia, Arduino, JavaScript and REST.
    
In this context, with Plotly we refer to the R package for to create interactive graphs.


### Advantages of using Plotly

- Seamless integration with R and `ggplot2`, through the `ggplotly()` function, that converts your static plots to an interactive web-based version!

- Diverse range of plots and customization (wide array of chart types with extensive customization options), as well as the possibility to zoom in and out into the plot, or select a certain part of the plot and explore it specifically.

- HTML web integration of graphs, makes visualization more accessible and intuitive. 

### Some visualization examples.

Many people are surprised that one is required to pay for using the public toilets in Berlin. Let's find out how far you would need to go to use a free toilet after a couple of beers with your friends.

<br>

Let's use this dataset for the public toilets in Berlin: [Berlin Public Toilet Data](https://www.kaggle.com/datasets/ryanjt/berlin-public-toilets-location/) database.
<br>
```{r}
#Load packages:
library(tidyverse)
library(plotly)
library(dash)

```

**Q.1.** Find out the district-wise distribution of public toilets across Berlin.

```{r}

#1. Prepare data
toilets <- read.csv("./data/berliner-toiletten-standorte.csv")

str(toilets) #tibble

toilets <- toilets |>
        mutate(Free = if_else(Price == 0, "free", "paid"))

#Filter the districts based on postal codes:

#Sources
#1. Wikipedia - https://simple.wikipedia.org/wiki/Postal_codes_in_Germany
#2. Cybo - https://postal-codes.cybo.com/germany/berlin/

toilets <- toilets |>
        mutate(District = case_when(
          (PostalCode >= 10115 & PostalCode <= 10179) | 
            (PostalCode >= 13347 &PostalCode <= 13359) ~ "Mitte",
          
          (PostalCode >= 10179 & PostalCode <= 10249) |
            (PostalCode >= 10961 & PostalCode <= 10999) ~ "Friedrichshain-Kreuzberg",
          
          (PostalCode >= 10367 & PostalCode <= 10407) | 
            (PostalCode >= 10318 & PostalCode <= 10319) ~ "Lichtenberg",
          
          (PostalCode >= 10405 & PostalCode <= 10439) | 
            (PostalCode >= 13086 & PostalCode <= 13129) ~ "Pankow",
          
          (PostalCode >= 10585 & PostalCode <= 10779) | 
            (PostalCode >= 12159 & PostalCode <= 12163) | 
            (PostalCode >= 13629 & PostalCode <= 14053) ~ "Charlottenburg-Wilmersdorf",
          
          (PostalCode >= 12099 & PostalCode <= 12109) | 
            (PostalCode >= 10777 & PostalCode <= 10783) | 
            (PostalCode >= 10823 & PostalCode <= 10829) | 
            (PostalCode >= 12305 & PostalCode <= 12309) | 
            (PostalCode >= 12277 & PostalCode <= 12279) ~ "Tempelhof-Schöneberg",
          
          (PostalCode >= 10999 & PostalCode <= 12059) | 
            (PostalCode >= 12555 & PostalCode <= 12559) | 
            (PostalCode >= 12487 & PostalCode <= 12489) ~ "Treptow-Köpenick",
          
          (PostalCode >= 12163 & PostalCode <= 12169) | 
            (PostalCode >= 14163 & PostalCode <= 14169) | 
            (PostalCode >= 12203 & PostalCode <= 12209) | 
            (PostalCode >= 12247 & PostalCode <= 12249) | 
                                  (PostalCode == 14109) ~ "Steglitz-Zehlendorf",
          
          (PostalCode >= 13407 & PostalCode <= 13439) | 
            (PostalCode >= 13505 & PostalCode <= 13509) ~ "Reinickendorf",
          
          (PostalCode >= 12043 & PostalCode <= 12059) ~ "Neukölln",
          
          (PostalCode >= 12679 & PostalCode <= 12689) ~ "Marzahn-Hellersdorf",
          
          (PostalCode >= 13599 & PostalCode <= 13629 ~ "Spandau")
      ))
```

```{r}
#Plot the distribution using plotly:


```

**Q.2.** Can you comment on the number of free toilets vs. paid public toilets across Berlin? Based on this dataset, which district in Berlin has the highest and lowest number of free public toilets?
<br>

**Q.3.** Using the 'beers' dataset, use ggplot to generate a parallel coordinates plot for the first 20 beers. Then transform the ggplot to a interactive plot using ggplotly.
(Hint: Use package "GGally" to generate a parallel coordinates plot. Beware of NA's!)

```{r}

```

# Dash
### What is Dash?

A web application framework for R (and Python and other programming languages) that allows creating interactive, web-based data dashboards (indeed), turning your interactive plots into full-blown web applications.

### Key features

- **HTML layout**: 
Design your dashboard layout using simple HTML components, without having to write any actual HTML. Indeed, it structure dashboard components using a set of pre-defined HTML tags within R.

- Interactivity through **callbacks**: 
callbacks define interactive behavior by specifying how input changes should modify output components. 
Essentially, when changing a dropdown or move a slider, a callback updates a chart, text, or any other component in real-time. 

- **Integration with Plotly** visualizations: Plotly charts can be easily included in Dash, making them part of an    
    interactive dashboard where each component can talk to others.

### Some visualization examples.

We will use the same toilets dataset.

```{r}

#0. Instantiate Dash App
app <- dash_app()
```

We mutate the table to add the columns for any specific filters that we require.

```{r}

toilets <- toilets |>
  mutate(Accessibility = if_else(isHandicappedAccessible == 0, "Handicapped Accessible", "Not Handicapped Accessible")) |>
  mutate(BabyFacilities = if_else(hasChangingTable == 0, "Changing table", "No changing table"))

```

We would like to see the distribution of the toilets across Berlin, based on whether they are paid or not.

```{r}
#2 Back-end controls

#Input component(s)

#We assign different colours to different districts.
color_options <- c(unique(toilets$District))

#We filter the toilets based on whether they are free or paid.
price_options <- c(unique(toilets$Free))
price_filter <- dccChecklist(
  id = 'price-filter',
  options = price_options,
  value = price_options #default, choose all
)

#We can also add additional filters like accessibility for the handicapped and child facilities:

handicap_options <- c(unique(toilets$Accessibility))
handicap_filter <- dccChecklist(
  id = 'handicap-filter',
  options = handicap_options,
  value = handicap_options #default, choose all
)

baby_options <- c(unique(toilets$BabyFacilities))
baby_filter <- dccChecklist(
  id = 'baby-filter',
  options = baby_options,
  value = baby_options #default, choose all
)
```

We assign the graph to a new variable and give instructions for callback based on the input(s) we provided.

```{r}

#Output component(s)
toilets_map <- dccGraph(id = 'toilets-map')

#Callback(s)
app |> add_callback(
  output('toilets-map', 'figure'),
  list(
    #input('district-dropdown', 'value'),
    input('price-filter', 'value'),
    input('handicap-filter', 'value'),
    input('baby-filter', 'value')
  ),
  function(price_input, handicap_input, baby_input) { #district_input
    #function(color_input) {
    toilets <- toilets |>
      filter(Free %in% price_input) |>
      filter(Accessibility %in% handicap_input) |>
      filter(BabyFacilities %in% baby_input)
    fig = toilets |>
      filter(District != "NA") |>
      plot_ly(lat = ~Latitude, lon = ~Longitude, type = "scattermapbox",
              split = ~District, size = 5, #toilets[[color_input]]
              hoverinfo = "text+name", hovertext = ~Street) |>
      layout(
        title = "Berlin public toilets",
        mapbox = list(
          style = 'open-street-map',
          zoom = 10,
          center = list(lat = ~median(Latitude), lon = ~median(Longitude)))
      )
    return(fig) #not needed but just to be explicit
  }
)
```

We also need to define the aesthetics of the dashboard:

```{r}

#3 Front-end layout

layout_elements <- list(
  h1("IDS Workshop - Plotly-Dash test"),
  div("Using the Berlin Public Toilets dataset to test Scatter Map functionality..."),
  br(),
  div("Type:"),
  price_filter,
  br(),
  div("Accessibility:"),
  handicap_filter,
  br(),
  div("Baby facilities:"),
  baby_filter,
  br(),
  toilets_map
)
```

Finally, we run the app:

```{r, eval=FALSE}

#4 Main app
app |> set_layout(layout_elements)
app |> run_app()
```

We can also add additional filters to the data, as and when required.

**Q.4.** Find the distribution of paid public toilets in Berlin, that are also handicapped accessible and have a changing table.

```{r}

```

**Q.5.** You have just submitted the materials for the IDS workshop on Monday and it is officially the end of the mid-term exam week. You go out with your friends to grab a few beers. And the weather is really cold. Are there any free toilets in Mitte that you could use after a grabbing some drinks with your friends?

# Thank you!

<br>